---
title: "Tooth Growth Analysis"
author: "Americo"
date: "19 luglio 2015"
output: html_document
---

##Utils
```{r loading packages}
require(dplyr)
require(ggplot2)
require(datasets)
require(tidyr)
options(digits = 3)
setwd("/Users/Americo/Documents/Education/Data_science/Coursera/Statinference/project/ass2_statinference")
```
##Goals
Now in the second portion of the class, we're going to analyze the ToothGrowth data in the R datasets package. 
Load the ToothGrowth data and perform some basic exploratory data analyses 
Provide a basic summary of the data.
Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
State your conclusions and the assumptions needed for your conclusions. 
Some criteria that you will be evaluated on
Did you  perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?
Did the student perform some relevant confidence intervals and/or tests?
Were the results of the tests and/or intervals interpreted in the context of the problem correctly? 
Did the student describe the assumptions needed for their conclusions?

##Dataset
Loading the dataset and studying the variables through the codebook and some initial exploration.
```{r loading dataset}
data(ToothGrowth)
?ToothGrowth
head(ToothGrowth)
tail(ToothGrowth)
dim(ToothGrowth)
str(ToothGrowth)
```

The dataset is composed by measurements on 10 pigs regarding the length of teeth (variable `len`) after the somministration of three different dose of vitamin (the variable `dose`) and two delivery methods (orange juice or ascorbic acid - the variable `supp`).
There is not a column dedicated to the ID of the ten pigs, so I assume that the measurements are ordered this way: each ten observations represents a dose of vitamin, the first 30 with a delivery methods and the second 30 with the other one.

##Exploratory data analysis
Esploro il dataset: le dimensioni, i missing values, le distribuzioni univariate, le statistiche di base (boxplot condizionati ai fattori, ecc)


##Data manipulation
I will try to check if a manipulation of the dataset can be useful for my analysis
```{r manipulation}
myToothGrowth <- ToothGrowth
mydfs <- list()
j <- 1:10
for (i in 1:6) {
        mydfs[[i]] <- myToothGrowth[j,]
        names(mydfs[[i]]) <- c(paste("len", i), paste("supp", i),paste("dose", i))
        j <- j + 10
}
myToothGrowth2 <- cbind(mydfs[[1]], mydfs[[2]],mydfs[[3]],mydfs[[4]],mydfs[[5]],mydfs[[6]])
myToothGrowth2$pig_id <- 1:10

```

##Statistical analysis


##Conclusions